---
title: "P8105 - Homework 2"
author: "Jake W. Coldiron - jwc2177 -"
date: "4 October 2023"
output: github_document
---

Before we begin, let's initialize a few helpful libraries since we're working with .csv files and .xlsx files.


```{r Initialize Libraries, collapse = TRUE}
#Initialize Libraries

library(tidyverse)

library(readxl)
```

# I. Problem 1

This problem uses the FiveThirtyEight data; these data were gathered to create the interactive graphic on this page. In particular, we’ll use the data in pols-month.csv, unemployment.csv, and snp.csv. Our goal is to merge these into a single data frame using year and month as keys across datasets.

## i. pols-month.csv

First, clean the data in pols-month.csv:

1. Use separate() to break up the variable mon into integer variables year, month, and day;
1. replace month number with month name; 
1. create a president variable taking values gop and dem, and remove prez_dem and prez_gop; 
1. and remove the day variable.

```{r pols_month piped and cleaned}
#pols_month piped and cleaned

pols_df = read_csv("./p8105_hw2_jwc2177_data/fivethirtyeight_datasets/pols-month.csv") %>% 
  
  separate(mon, into = c("year", "month"), sep = 4) %>% 
  
  separate(month, into = c("month", "day"), sep = 4) %>% 
  
  mutate(month = recode(month, "-01-" = "january", "-02-" = "february", "-03-" = "march", "-04-" = "april", "-05-" = "may", "-06-" = "june", "-07-" = "july", "-08-" = "august", "-09-" = "september", "-10-" = "october", "-11-" = "november", "-12-" = "december")) %>% 
  
  pivot_longer(
    c("prez_gop", "prez_dem"),
    names_to = "prez",
    values_to = "whitehouse"
    ) %>% 
  
  mutate(prez = recode(prez, "prez_gop" = "gop", "prez_dem" = "dem")) %>% 
  
#Apparently there was a value of 2 for the President in September 1974, October 1974, November 1974, and December 1974. This does not make sense as the US only has one President at a time. 
  
#The value is likely referring to the resignation of Richard Nixon. Immediately after Nixon resigned on 9 August 1974, his  Vice President Gerald Ford was sworn in as President. If there was going to be a value of 2 for the President, then August 1974 would make the most sense, however there is no value of 2 in August 1974 only the aforementioned months. 
  
#This doesn't make much sense. Regardless, the data was processed as given.
  
  mutate(whitehouse = recode(whitehouse, "0" = "no", "1" = "yes", "2" = "yes")) %>% 
  
  select(-day) %>% 
  
  relocate(year, month, prez, whitehouse, gov_dem, gov_gop, rep_dem, rep_gop, sen_dem, sen_gop)

pols_df
```

## ii. snp.csv

Second, clean the data in snp.csv using a similar process to the above. For consistency across datasets:

1. arrange according to year and month, 
1. and organize so that year and month are the leading columns.

```{r snp piped and cleaned}
#snp piped and cleaned

snp_df = read_csv("./p8105_hw2_jwc2177_data/fivethirtyeight_datasets/snp.csv") %>%
  separate(date, into = c("month", "day", "year")) %>% 
  
  mutate(month = recode(month, "1" = "january", "2" = "february","3" = "march","4" = "april","5" = "may","6" = "june","7" = "july","8" = "august","9" = "september","10" = "october","11" = "november","12" = "december")) %>% 
  
  mutate(year = paste("20", year, sep = "")) %>% 
  
  mutate(year = as.numeric(year)) %>% 
  
  mutate(year = ifelse(year > 2015, year - 100, year)) %>% 
  
  select(-day) %>% 
  
  relocate(year, month, close) %>% 
  
  rename(snp_close = close)

snp_df
```


## iii. unemployment.csv

Third, tidy the unemployment data so that it can be merged with the previous datasets. This process will involve switching from “wide” to “long” format; ensuring that key variables have the same name; and ensuring that key variables take the same values.

```{r unemployment piped and cleaned}
#unemployment piped and cleaned

unemployment_df = read_csv("./p8105_hw2_jwc2177_data/fivethirtyeight_datasets/unemployment.csv") %>%
  
  janitor::clean_names() %>% 
  
  pivot_longer(
    jan:dec,
    names_to = "month",
    values_to = "unemployment_percent"
  ) %>% 
  
  mutate(month = recode(month, "jan" = "january", "feb" = "february", "mar" = "march", "apr" = "april", "may" = "may", "jun" = "june", "jul" = "july", "aug" = "august", "sep" = "september", "oct" = "october", "nov" = "november", "dec" = "december"))
  

unemployment_df

```

```{r, eval = false}
#unemployment piped and cleaned

unemployment_df_test = read_csv("./p8105_hw2_jwc2177_data/fivethirtyeight_datasets/unemployment.csv") %>% 
  
  janitor::clean_names()

```

## iv. Joining the Datasets 

Fourth, join the datasets by merging snp into pols, and merging unemployment into the result.


## v. Describing the Joined Dataset 

Fifth, write a short paragraph about these datasets. Explain briefly what each dataset contained, and describe the resulting dataset (e.g. give the dimension, range of years, and names of key variables).
